Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-04-24 manuscript/main.qmd
Update intro

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
---

@@

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes ($\delta^18^O$), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales (e.g. REF), recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most fundamental level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specifics regions (e.g. REFs). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^{18}O$, $\Delta_{47}$, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised $\delta^18^O$ samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This resulted in most \$\delta^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@Hollis2019].

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g. $\delta^18^O$) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted by their mean temperature values sampled from their respective normal distributions.
:::

@@

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMsST. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies. Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

>>>>

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
author:
  - "---"
  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
  - "---"
  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
  - "---"
  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
abstract: \newpage Abstract
format:
  docx:
    reference-doc: template.docx
    highlight-style: github
    fig-dpi: 900
  pdf:
    documentclass: scrartcl
    papersize: letter
    geometry:
      - top=25.4mm
      - bottom=25.4mm
      - left=25.4mm
      - right=25.4mm
      - heightrounded
bibliography: references.bib
csl: climate-of-the-past.csl
editor: source
---

@@

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This resulted in most &delta;^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@Hollis2019].

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g. &delta;^18^O) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted by their mean temperature values sampled from their respective normal distributions.
:::

@@

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMsST. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies. Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
